An Efficient Forecasting Approach for the Real-Time Reduction of Boundary Effects in Time-Frequency Representations

### 2020-07-18 Notes/BoundEffRed.tex
code and typos
typos

Windowing is a common method for performing local analysis. Among many others, STFT~\cite{Flandrin:1999}, continuous wavelet transform (CWT)~\cite{Da1992}, synchrosqueezing transform (SST) \cite{Daubechies11synchrosqueezed}, and reassignment~\cite{Auger13time} (RS) are representations that fall back on the use of an analysis window. Let $x:I\to\RR$ denote the observed signal, where $I$ denotes the finite interval where the signal is measured. Let $g_s:\RR\to\RR$ denote the analysis window, where $s$ is a shape parameter. The support of $g_s$ is localized around the origin and is small with respect to $|I|$. The translation operator is $T_\tau$ defined as:
\[
T_\tau f = f(t-\tau)\ ,\quad \forall\ f:\RR\to\RR\ .
\]
Then, the local analysis of $x$ around the instant $\tau\in I$ rely on the evaluation of the following inner product:
\begin{equation}
V_x(s,\tau) = \langle x, T_\tau g_s \rangle_I \ .
\label{eq:windowing}
\end{equation}
A major shortcoming of this technique occurs when analyzing the signal $x$ {\em near the boundaries} of the interval $I$. Clearly, at these points, half of the information is missing. Consequently, the results of the inner product~\eqref{eq:windowing} are distorted. This phenomenon is usually understood as the \emph{boundary effect}. We display on the bottom left of Fig.~\ref{fig:ex.intro} the result of the SST of a PPG (see section~\ref{ssse:ppg} for a comprehensive description). The distortion resulting from boundary effects is clearly visible on the right side of this representation. Indeed, while in the major left part of the image, clear lines stand out, they become blurred as they approach the right boundary of the image. Therefore, estimations of signal characteristics, like instantaneous frequencies~\cite{Delprat92asymptotic} or amplitudes, from this TF representation appear to be imprecisely determinable (or even likely to fail) in the vicinity of the boundaries. Moreover, this boundary effect would unavoidably limit applying these window-based TF analysis tools for real-time analysis purposes. It is thus desirable to have a solution to eliminating the boundary effects.

@@

The paper is organized in the following. In section~\ref{se:algo}, we provide an extension method based on a linear dynamic model. We derive the corresponding algorithm for boundary effects reduction. In section~\ref{se:theoretical}, we show that the dynamic model we consider is sufficient to extend signals takings the form of sums of sine waves. An evaluation of the theoretical performance of our algorithm on a class of signals, the sums of sine waves, is given in section~\ref{se:theoretical}. In section~\ref{se:results}, we compare our extension method with more sophisticated methods such as EDMD, GPR, or TBATS. We show that our algorithm gives fast results of reasonable quality. Finally, we evaluate the performance of our boundary effects reduction algorithm on biomedical signals, such as respiratory signals, and compare it to the theoretical results.

@@

Various improvements can be considered to make the algorithm more robust. In particular, we have noticed (see Fig.~\ref{fig:THO.failure}) that when the regular oscillations of the observed signal break, the forecasting step is no longer relevant, and only slows down the calculation of the TF representation. A preliminary step should then be added to the algorithm to detect signal activity, and disable the forecasting step when possible. A more fundamental improvement would be to perform the forecasting step directly in the TF plane, to specifically predict the evolution of each instantaneous frequency. This would lead to seeing a TF representation as a multivariate time series (corresponding to each frequency band). We will explore these possibilities in our future work.

>>>>

Windowing is a common method for performing local analysis. Among many others, STFT~\cite{Flandrin:1999}, continuous wavelet transform (CWT)~\cite{Da1992}, synchrosqueezing transform (SST) \cite{Daubechies11synchrosqueezed}, and reassignment~\cite{Auger13time} (RS) are representations that fall back on the use of an analysis window. Let $x:I\to\RR$ denote the observed signal, where $I$ denotes the finite interval where the signal is measured. Let $g_s:\RR\to\RR$ denote the analysis window, where $s$ is a shape parameter. The support of $g_s$ is localized around the origin and is small with respect to $|I|$. The translation operator is $T_\tau$ defined as:
\[
T_\tau f = f(t-\tau)\ ,\quad \forall\ f:\RR\to\RR\ .
\]
Then, the local analysis of $x$ around the instant $\tau\in I$ rely on the evaluation of the following inner product:
\begin{equation}
V_x(s,\tau) = \langle x, T_\tau g_s \rangle_I \ .
\label{eq:windowing}
\end{equation}
A major shortcoming of this technique occurs when analyzing the signal $x$ {\em near the boundaries} of the interval $I$. Clearly, at these points, half of the information is missing. Consequently, the results of the inner product~\eqref{eq:windowing} are distorted. This phenomenon is usually understood as the \emph{boundary effect}. The result of the SST of a PPG is shown in the lower left corner of Fig.~\ref{fig:ex.intro} (see section~\ref{ssse:ppg} for a comprehensive description). The distortion resulting from boundary effects is clearly visible on the right side of this representation. Indeed, while in the major left part of the image, clear lines stand out, they become blurred as they approach the right boundary of the image. Therefore, estimations of signal characteristics, like instantaneous frequencies~\cite{Delprat92asymptotic} or amplitudes, from this TF representation appear to be imprecisely determinable (or even likely to fail) in the vicinity of the boundaries. Moreover, this boundary effect would unavoidably limit applying these window-based TF analysis tools for real-time analysis purposes. It is thus desirable to have a solution to eliminating the boundary effects.

@@

The paper is organized in the following way. In section~\ref{se:algo}, we provide an extension method based on a linear dynamic model. We derive the corresponding algorithm for boundary effects reduction. In section~\ref{se:theoretical}, we show that the dynamic model we consider is sufficient to extend signals takings the form of sums of sine waves. An evaluation of the theoretical performance of our algorithm on a class of signals, the sums of sine waves, is given in section~\ref{se:theoretical}. In section~\ref{se:results}, we compare our extension method with more sophisticated methods such as EDMD, GPR, or TBATS. We show that our algorithm gives fast results of reasonable quality. Finally, we evaluate the performance of our boundary effects reduction algorithm on biomedical signals, such as respiratory signals, and compare it to the theoretical results.

@@

Various improvements can be considered to make the algorithm more robust. In particular, we have noticed (see Fig.~\ref{fig:THO.failure}) that when the regular oscillations of the observed signal break, the forecasting step is no longer relevant, and only slows down the calculation of the TF representation. A preliminary step should then be added to the algorithm to detect signal activity, and disable the forecasting step when possible. A more fundamental improvement would be to perform the forecasting step directly in the TF plane, to specifically predict the evolution of each instantaneous frequency. This would lead to seeing a TF representation as a multivariate time series, with each frequency band containing a univariate time series. We will explore these possibilities in our future work.

### 2020-07-18 Notes/results.tex
precision
typos

\paragraph{Influence of the training dataset size $K$} Here, the noise variance $\sigma$ is set to $\sigma=10^{-2}$. Then, the forecasting algorithm is run on 3000 realizations on the discrete signal $\bx$ for three different values of $K$, logarithmically equispaced from $4.5\times 10^{2}$ to $2\times 10^{3}$. For each of these values, we determine the experimental bias $\mu_{\xp}[N-1+\ell]$ and variance $\gamma_{\xp}[N-1+\ell,N-1+\ell]$ in function of the forecasting sample index $\ell$ (going from $1$ to $500$).

As in the previous study, the experimental bias vanishes when $K$ increases, confirming the approximation result~\eqref{eq:mean.error}. Besides, the experimental variance  is displayed on Fig.~\ref{fig:res.size.sine}, and compared with the associated theoretical variance~\eqref{eq:cov.error.2}. Each color corresponds to these experimental results obtained for a given value of $K$. This result validates the asymptotic behavior provided by~\eqref{eq:cov.error.2}, and we can show that the second-order moment $\gamma[\ell,\ell]$ is less dependent on $K$ when $K$ is large.

@@

\subsubsection{Adaptive harmonic model}
\label{ssse:res.ahm}
We now consider a signal satisfying the AHM so that the instantaneous frequencies and amplitudes of its components vary from time to time. The deterministic component $\bz$ of the random vector $\bx$ (constructed following the model~\eqref{eq:model.noise}) takes the following form, for all $n\in\{1,\ldots,N\}$:
\[
\bx[n] = \cos\left(2\pi \phi_1[n] \right) + R[n]\cos\left(2\pi \phi_2[n] \right) \ ,
\] 
where the instantaneous amplitude $R$ is given by:
\[
R[n] = 1.4 + 0.2\cos\left(4\pi\frac{n}{N}\right)\ ,
\]
and the instantaneous phases are such that:
\begin{align*}
\phi_1[n] &= \frac{p_1}{M}\left( n + \frac{0.01}{2\pi}\cos\left(2\pi\frac{n}{N}\right) \right) \\[-1mm]
\phi_2[n] & = p_2\frac{n}{M} + \frac{20}{2N\fs}n^2
\end{align*}
Besides, the noise is chosen to be Gaussian: $\bw\sim\cN(\bzero,\bI)$, and we take: $N=10^4$, $M=750$, $p_1=10$, $p_2=23$.

@@

We then apply {\sf BoundEffRed} (\ie~Algorithm~\ref{alg:boundary}) for diverse TF representations: STFT, SST, RS, as well as \textit{Concentration of Frequency and Time} (ConceFT), a generalized multitaper SST-based representation introduced in~\cite{Daubechies16conceft}. In Table~\ref{tab:THO}, we give the averaged performance index~\eqref{eq:index.perf}, evaluated the whole TF representations (including boundaries). Even though {\sf SigExt} performs somehow moderately, the boundary effects are dramatically reduced on the TF representations, in the same order of magnitude than with the forecastings given by EDMD or GPR. Notice that the extension length $L$ has been set accordingly to the window length used by the TF analysis tool. For instance, the window length used to evaluate the STFT is of $1500$ samples. To prevent the STFT from being sensitive to the boundaries, we set $L=750$. In this way, the evaluation of the spectral content of the signal near its boundaries is not limited by a lack of information all along the window support. From now on, all results are given for $L$ equal to the half of the width of the window used in the TF transform.

>>>>

\paragraph{Influence of the training dataset size $K$} Here, the noise variance $\sigma$ is set to $\sigma=10^{-2}$. Then, the forecasting algorithm is run on 3000 realizations on the discrete signal $\bx$ for three different values of $K$, logarithmically equispaced from $4.5\times 10^{2}$ to $2\times 10^{3}$. For each of these values, we determine the experimental bias $\mu_{\xp}[N-1+\ell]$ and variance $\gamma_{\xp}[N-1+\ell,N-1+\ell]$ in function of the forecasting sample index $\ell$, going from $1$ to $500$.

As in the previous study, the experimental bias vanishes when $K$ increases, confirming the approximation result~\eqref{eq:mean.error}. Besides, the experimental variance  is displayed in Fig.~\ref{fig:res.size.sine}, and compared with the associated theoretical variance~\eqref{eq:cov.error.2}. Each color corresponds to these experimental results obtained for a given value of $K$. This result validates the asymptotic behavior provided by~\eqref{eq:cov.error.2}, and we can show that the second-order moment $\gamma[\ell,\ell]$ is less dependent on $K$ when $K$ is large.

@@

\subsubsection{Adaptive harmonic model}
\label{ssse:res.ahm}
We now consider a signal satisfying the AHM so that the instantaneous frequencies and amplitudes of its components vary from time to time. The deterministic component $\bz$ of the random vector $\bx$, constructed following the model~\eqref{eq:model.noise}, takes the following form, for all $n\in\{1,\ldots,N\}$:
\[
\bx[n] = \cos\left(2\pi \phi_1[n] \right) + R[n]\cos\left(2\pi \phi_2[n] \right) \ ,
\] 
where the instantaneous amplitude $R$ is given by:
\[
R[n] = 1.4 + 0.2\cos\left(4\pi\frac{n}{N}\right)\ ,
\]
and the instantaneous phases are such that:
\begin{align*}
\phi_1[n] &= \frac{p_1}{M}\left( n + \frac{0.01}{2\pi}\cos\left(2\pi\frac{n}{N}\right) \right) \\[-1mm]
\phi_2[n] & = p_2\frac{n}{M} + \frac{20}{2N\fs}n^2
\end{align*}
Besides, the noise is chosen to be Gaussian: $\bw\sim\cN(\bzero,\bI)$, and we take: $N=10^4$, $M=750$, $p_1=10$, $p_2=23$.

@@

We then apply {\sf BoundEffRed} (\ie~Algorithm~\ref{alg:boundary}) for diverse TF representations: STFT, SST, RS, as well as \textit{Concentration of Frequency and Time} (ConceFT), a generalized multitaper SST-based representation introduced in~\cite{Daubechies16conceft}. In this application and throughout the document, we use a Gaussian window as the analysis window, regardless of the representation involved. In Table~\ref{tab:THO}, we give the averaged performance index~\eqref{eq:index.perf}, evaluated the whole TF representations (including boundaries). Even though {\sf SigExt} performs somehow moderately, the boundary effects are dramatically reduced on the TF representations, in the same order of magnitude than with the forecastings given by EDMD or GPR. Notice that the extension length $L$ has been set accordingly to the window length used by the TF analysis tool. For instance, the window length used to evaluate the STFT is of $1500$ samples. To prevent the STFT from being sensitive to the boundaries, we set $L=750$. In this way, the evaluation of the spectral content of the signal near its boundaries is not limited by a lack of information all along the window support. From now on, all results are given for $L$ equal to the half of the width of the window used in the TF transform.
